Staged Abstract Interpreters

### 2019-09-04 paper/40_absinterp.tex
space

\begin{figure}[h!]
  \centering
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => for {
    ρ <- ask_env; σ <- get_store; in <- ask_in_cache; out <- get_out_cache; val cfg = (e, ρ, σ)
    rt <- if (out.contains(cfg)) for { // ask if out already contains the desired result
            (v, σ) <- lift_nd[(Value, Store)](out(cfg))
             _ <- put_store(σ)
          } yield v
          else for {
            _ <- put_out_cache(out + (cfg → in.getOrElse(cfg, ⊥)))
            v <- ev(fix(ev))(e)
            σ <- get_store
            _ <- update_out_cache(cfg, (v, σ))
          } yield v
  } yield rt
\end{lstlisting}
\vspace{-1em}
\caption{The unstaged co-inductive caching algorithm.}
\label{fig:coind_cache}
\end{figure}

>>>>

\begin{figure}[h!]
  \centering
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => for {
    ρ <- ask_env; σ <- get_store; in <- ask_in_cache; out <- get_out_cache; val cfg = (e, ρ, σ)
    rt <- if (out.contains(cfg)) for { // ask if out already contains the desired result
            (v, σ) <- lift_nd[(Value, Store)](out(cfg));  _ <- put_store(σ)
          } yield v
          else for {
            _ <- put_out_cache(out + (cfg → in.getOrElse(cfg, ⊥)))
            v <- ev(fix(ev))(e);  σ <- get_store;  _ <- update_out_cache(cfg, (v, σ))
          } yield v
  } yield rt
\end{lstlisting}
\vspace{-1em}
\caption{The unstaged co-inductive caching algorithm.}
\label{fig:coind_cache}
\vspace{-1em}
\end{figure}

### 2019-09-04 paper/50_stageabs.tex
misc
space

\begin{figure}[h!]
  \centering
\begin{lstlisting}
  def fix_cache(e: Expr): Ans = for {
    ρ   <- ask_env;  σ <- get_store;  in <- ask_in_cache;  out <- get_out_cache
    cfg <- lift_nd[Config](Set((unit(e), ρ, σ)))
    res <- lift_nd[(Set[ValSt], Cache)](Set(
      // a next-stage value of type Rep[(Set[ValSt], Cache)] generated by if
      if (out.contains(cfg)) (repMapToMapOps(out).apply(cfg).toList, out)
      else { val m: Ans = for {
               _ <- put_out_cache(out + (cfg → in.getOrElse(cfg, ⊥)))
               v <- eval(fix_cache)(e)
               σ <- get_store
               _ <- update_out_cache(cfg, (v, σ))
             } yield v
             m(ρ)(σ)(in)(out) }))
    _ <- put_out_cache(res._2);  vs <- lift_nd(res._1);  _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}
\caption{The staged co-inductive caching algorithm.}
\label{fig:staged_coind_cache}
\end{figure}

@@

@@

Although every component except @Expr@s are staged, we treat the data
structures such as @Map@s as black-boxes, which means the operations on a @Map@
directly become code in the next stage, but we don't inspect any further inside.
As we identified when introducing the generic interface, the keys of @Env@s are
identifiers (i.e., strings) in the program, which are completely known
statically. This leaves us a chance to further specialize the data structures.
For example, let's assume that the @Map[K, V]@ is implemented as a hash map. If
the keys of type @K@ are all known statically, then the indices also can be
computed statically. Thus the specialized map would be an array of type
@Array[Rep[V]]@, whose elements are next-stage values; the size of the array is
also known statically. An access to the map is translated into an access to the
array with pre-determined index during staging.

>>>>

@@

\begin{figure}[h!]
\centering
\begin{lstlisting}
  def fix_cache(e: Expr): Ans = for {
    ρ   <- ask_env;  σ <- get_store;  in <- ask_in_cache;  out <- get_out_cache
    cfg <- lift_nd[Config](Set((unit(e), ρ, σ)))
    res <- lift_nd[(Set[ValSt], Cache)](Set(
      // a next-stage value of type Rep[(Set[ValSt], Cache)] generated by if
      if (out.contains(cfg)) (repMapToMapOps(out).apply(cfg).toList, out)
      else { val m: Ans = for {
               _ <- put_out_cache(out + (cfg → in.getOrElse(cfg, ⊥)))
               v <- eval(fix_cache)(e);  σ <- get_store;  _ <- update_out_cache(cfg, (v, σ))
             } yield v
             m(ρ)(σ)(in)(out) }))
    _ <- put_out_cache(res._2);  vs <- lift_nd(res._1);  _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}
\caption{The staged co-inductive caching algorithm.}
\label{fig:staged_coind_cache}
\vspace{-2em}
\end{figure}

@@

Although every component except @Expr@s are staged, we treat the data
structures such as @Map@s as black-boxes, which means the operations on a @Map@
directly become code in the next stage, but we don't inspect any further inside.
As we identified when introducing the generic interface, the keys of @Env@s are
identifiers (i.e., strings) in the program, which are completely known
statically. This leaves us a chance to further specialize the data structures.
For example, let us assume that the @Map[K, V]@ is implemented as a hash map. If
the keys of type @K@ are all known statically, then the indices also can be
computed statically. Thus the specialized map would be an array of type
@Array[Rep[V]]@, whose elements are next-stage values; the size of the array is
also known statically. An access to the map is translated into an access to the
array with pre-determined index during staging.

### 2019-09-04 paper/abstract.tex
space

In this paper, we present a unifying framework that naturally extends the first
Futamura projection from concrete interpreters to abstract interpreters. Our
approach derives a sound staged abstract interpreter based on a
semantic-agnostic interpreter with type-level binding-time abstraction and
monadic abstraction. By using different instantiations of these abstractions,
the generic interpreter can flexibly behave in four modes: unstaged concrete
interpreter, staged concrete interpreter, unstaged abstract interpreter, or
staged abstract interpreter. As an example of \emph{abstraction without regret},
the overhead of these abstraction layers is eliminated in the generated code
after staging. We show that staging abstract interpreters is practical and
useful to optimize static analysis, while requiring less engineering effort,
without compromising soundness. We conduct three case studies, including a
comparison with \citeauthor{Boucher:1996:ACN:647473.727587}'s abstract
compilation, applications to various control-flow analyses, and a demonstration
that can be used for modular analysis. We also empirically evaluate the effect
of staging on the execution time. The experiment shows an order of magnitude
speedup with staging for control-flow analyses.

>>>>

In this paper, we present a unifying framework that naturally extends the first
Futamura projection from concrete interpreters to abstract interpreters. Our
approach derives a sound staged abstract interpreter based on a
semantic-agnostic interpreter with type-level binding-time abstraction and
monadic abstraction. By using different instantiations of these abstractions,
the generic interpreter can flexibly behave in four modes: unstaged concrete
interpreter, staged concrete interpreter, unstaged abstract interpreter, or
staged abstract interpreter. As an example of \emph{abstraction without regret},
the overhead of these abstraction layers is eliminated in the generated code
after staging. We show that staging abstract interpreters is practical and
useful to optimize static analysis, while requiring less engineering effort,
without compromising soundness. We conduct three case studies, including a
comparison with Boucher and Feeley's abstract compilation, applications to
various control-flow analyses, and a demonstration that can be used for modular
analysis. We also empirically evaluate the effect of staging on the execution
time. The experiment shows an order of magnitude speedup with staging for
control-flow analyses.
